Matched Source Waveform Inversion for Transmitted Waves

### 2024-02-28 repro/papers/awinum/paper.tex
more

\section{Introduction}
Matched Source Waveform Inversion (MSWI) is a variant of Full Waveform
Inversion (FWI) \cite[]{VirieuxOperto:09}, that sometimes overcomes
one of FWI's impediments, namely its tendency to stagnate at
suboptimal model estimates.

(...)
,
demonstrably far from global optima in controlled settings and often
uninformative of material structure in field use. It is possible that
such local descent algorithms are trapped in regions around local,
\section{Summary}
The version of MSWI discussed here uses acoustic wave propagation with
isotropic point sources and receivers.
The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
The model vectors $m=(\kappa,\rho)$ make up the domain of the forward
map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf
  x}_s)\}$, for specified source and receiver positions ${\bf x}_s, {\bf x}_r$ and
recording interval $[0,t_d]$.

@@

The approach to local optimization taken here (and in most work on FWI
and related topics) is based on the gradient of the objective defined
in equiation \ref{eqn:fwi}:
\begin{equation}
  \label{eqn:fwigrad}
  g = \nabla  J_{\rm FWI}[m;d] = DF[m]^T(F[m]w-d).
\end{equation}
In this formula, $DF[m]$ is the derivative of $F$ at $m$. 
This is the Euclidean (or $L^2$) gradient, that is, the vector $g$ for
which the Euclidean inner product
\begin{equation}
  \label{eqn:eucip}
  \langle g, \delta m\rangle = g^T\delta m
\end{equation}
with any other vector $\delta m$ gives the
rate of change in the direction of that vector of $J_{\rm FWI}$ at m.

For optimization within a set of slowly varying models (on the
wavelength scale),  it is appropriate to penalize oscillation of the
search vector. A convenient way to accompish this goal is the use of a
{\em weighted inner product} to define the gradient, rather than the
Euclidean inner product. A weight operator $W$ should be symmetric and
positive definite: then
\begin{equation}
  \label{eqn:wip}
  \langle g, \delta m\rangle_W = g^TW\delta m
\end{equation}
defines an alternative inner product. Comparing the definitions
\ref{eqn:fwigrad}, \ref{eqn:eucip}, and \ref{eqn:wip}, clearly the
vector $g_W$ for which $\langle g_W, \delta m \rangle_W$ gives the
rate of change of $J_{\rm FWI}$ at $m$ in the direction $\delta m$ is
\begin{equation}
  \label{eqn:fwiwgrad}
  g_W = W^{-1}\nabla  J_{\rm FWI}[m;d] =W^{-1} DF[m]^T(F[m]w-d).
\end{equation}

@@

It is possible to make the error between filtered predicted data and
observed data as small as one likes by choosing an appropriate filter
field $u$, so this error by itself is useless for estimating the
model. If $u(\bx_s,\bx_r,t)=\delta(t)$, on the other hand the filtered
predicted data is identical to the predicted data.
Therefore, some penalty for divergence of the filter $u$ from
$\delta(t)$ needs to supplement the filtered prediction error.. The works referenced in the Introduction mostly use the
mean-square of the filter scaled by $t$, and add it to the mean-square
of the filtered prediction error. This sum is the MSWI objective function:
\begin{equation}
  \label{eqn:filtpen}
  J_{\alpha,\sigma}[m,u;d]=\frac{1}{2}(\|K[u]F[m]w-d\|^2
  +\alpha^2\|tu\|^2 + \sigma^2\|u\|^2).
\end{equation}

@@

While the reduction in the (weighted) gradient norm indicates progress
towards a stationary point, it is not possible to claim that the final
estimate is in the vicinity of a local minimizer. The fit to data
obtained is not at all satisfactory. More discussion of this result
will be found later in the paper.

>>>>

\section{Introduction}
Matched Source Waveform Inversion (MSWI) is a variant of Full Waveform
Inversion (FWI) \cite[]{VirieuxOperto:09}, that sometimes overcomes
one of FWI's impediments, namely its tendency to stagnate at
suboptimal model estimates (``cycle-skip''), demonstrably far from global optima in
controlled settings and often uninformative of material structure in
field use. It is possible that such local descent algorithms are
trapped in regions around local minima, though such trapping, or even
the existance of local, non-global minima, are seldom established with
any rigor. MSWI
loosens the bond between predicted and observed data by interposing a
filter, adapted to map one to the other trace-by-trace, and penalizes
deviation of the filter from the Dirac delta. This soft penalty
effectively relaxes the FWI data-fitting problem and permits local
optimization methods to approximate data kinematics.

The aim of this paper is to review the theoretical basis of MSWI,
describe a practical computational framework for MSWI, and use it to
solve a synthetic acoustic inversion problem for which FWI fails. In
contrast, MSWI starting at the same initial estimate delivers a
improved velocity from which FWI succeeds in approaching the global
minimizer. The example is representative of a case (single arrival or
simple wavefront transmitted wave data) in which the MSWI objective
function is demonstrably close to mean square travel time error. Since
travel time inversion is generally not subject to cycle-skipping,
successful inversion is expected, and that is what we observe.

MSWI in the form described here was introduced by
\cite{HuangSymes2015SEG,HuangSymes:Geo17}. It is mathematically
equivalent to an {\em extended} formulation of the inverse problem, in
which acoustic point sources are allowed to depend on the receiving
sensors - a non-physical expansion of the simulation domain. A number
of other variants of FWI have been based on essentially the same
extension of acoustic modeling
\cite[]{Song:94c,Symes:94c,Plessix:00,LuoSava:11,LiAlkhalifah:21}. In
particular it is the basis of Adaptive Waveform Inversion (AWI) \cite[]{Warner:16,
  GuaschWarnerRavaut:GEO19,Warneretal:SEG21, Guaschetal:NPJDM20}. MSWI
is closely related to AWI, but not identical: AWI includes a
normalization of the adaptive filter, which makes the AWI objective
function an even closer approximation to travel time mean square error
than is MSWI, at least for transmitted wave data with a single
arriving wavefront \cite[]{Symes:24a}.

Transmission data with a single arrival is not just a case in which
MSWI and AWI are known to closely approximate travel time inversion:
it is the only case in which this link occurs. In particular, the link
is broken for transmission data exhibiting multiple arrivals: in the
presence of complex wavefronts, methods based on the source-receiver
extension are no less likely than FWI to cycle-skip
\cite[]{Symes:94c,HuangSymes:Geo17,Symes:24a}. However, the
source-receiver extension is not the only possible route to FWI
modification by artificially expanding the definition of energy
source. \cite{HuangNammourSymesDollizal:SEG19} overview modifications
of FWI based on various source extensions; some more recent advances
are described by
\cite{MetivierBrossier:SEG20,PladysBrossierLiMetivier:GEO21,LiAlkhalifah:21,Yongetal:GJI23,Opertoetal:GEO23}.
Numerical examples suggest that some of these extensions may avoid
cycle-skipping for transmitted wave data with complex wavefronts.

The next section describes versions of FWI and MSWI based on
acoustic wave propagation. The theoretical connection between MSWI and
travel time inversion is reviewed. The computations required to apply local
optimization to the objective functions of these inversion methods are
detailed, as is the variable projection reduction of MSWI and the use
of weighted norms in the domain of the simulation operator (wave
velocity or bulk modulus fields, in the setting developed here). The
third section presents an example, along with a detailed description
of the numerical methods for simulation and optimization methods used
in treating this example. The final section discusses some of the many
unresolved questions around MSWI and similar approaches to inverse
problems in wave propagation.

\section{Background}
The version of MSWI discussed here uses acoustic wave propagation with
isotropic point sources and receivers.
The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
The model vectors $m=(\kappa,\rho)$ make up the domain of the forward
map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf
  x}_s)\}$, for specified source and receiver positions ${\bf x}_s, {\bf x}_r$ and
recording interval $[0,t_d]$.

In the discussion that follows, the buoyancy $\beta$ will be regarded
as a fixed parameter. Thus $m$ is effectively the bulk modulus field $\kappa$.

@@

The approach to local optimization taken here (and in most work on FWI
and related topics) is based on the gradient of the objective defined
in equiation \ref{eqn:fwi}:
\begin{equation}
  \label{eqn:fwigrad}
  g = \nabla  J_{\rm FWI}[m;d] = D_m(F[m]w)^T(F[m]w-d).
\end{equation}
In this formula, $D_m(F[m]w)$ is the derivative of $F[m]w$ with
respect to $m$. 
This is the Euclidean (or $L^2$) gradient, that is, the vector $g$ for
which the Euclidean inner product
\begin{equation}
  \label{eqn:eucip}
  \langle g, \delta m\rangle = g^T\delta m
\end{equation}
with any other vector $\delta m$ gives the
rate of change in the direction of that vector of $J_{\rm FWI}$ at m.

For optimization within a set of slowly varying models (on the
wavelength scale),  it is appropriate to penalize oscillation of the
search vector. A convenient way to accompish this goal is the use of a
{\em weighted inner product} to define the gradient, rather than the
Euclidean inner product. A weight operator $W$ should be symmetric and
positive definite: then
\begin{equation}
  \label{eqn:wip}
  \langle g, \delta m\rangle_W = g^TW\delta m
\end{equation}
defines an alternative inner product. Comparing the definitions
\ref{eqn:fwigrad}, \ref{eqn:eucip}, and \ref{eqn:wip}, clearly the
vector $g_W$ for which $\langle g_W, \delta m \rangle_W$ gives the
rate of change of $J_{\rm FWI}$ at $m$ in the direction $\delta m$ is
\begin{equation}
  \label{eqn:fwiwgrad}
  g_W = W^{-1}\nabla  J_{\rm FWI}[m;d] =W^{-1} D_m(F[m]w)^T(F[m]w-d).
\end{equation}

@@

Note that the filtered predicted data may be viewed as the predicted
data for an {\em extended source}, with the source wavelet at location
$\bx_s$ replaced by $u(\bx_s,\bx_r,\cdot) * w(\bx_s,\cdot)$. That is,
the adaptive filter construction is equivalent to allowing the source
to depend on receiver position as well as source position - one source
wavelet for each source-receiver pair. This is an extension of
standard modeling, in that the domain (bulk modulus buoyancy, wavelet)
is larger than in the conventional formulation, and coincides with it
under the special condition that the source wavelet is independent of
the receiver location. This  {\em
  source-receiver extension}\cite[]{HuangSymes2015SEG} is a key ingredient in a number of
other papers on modifications of FWI, as mentioned in the Introduction.

It is possible to make the error between filtered predicted data and
observed data as small as one likes by choosing an appropriate filter
field $u$, so this error by itself is useless for estimating the
model. If $u(\bx_s,\bx_r,t)=\delta(t)$, on the other hand the filtered
predicted data is identical to the predicted data.
Therefore, some penalty for divergence of the filter $u$ from
$\delta(t)$ needs to supplement the filtered prediction error.. The works referenced in the Introduction mostly use the
mean-square of the filter scaled by $t$, and add it to the mean-square
of the filtered prediction error. This sum is the MSWI objective function:
\begin{equation}
  \label{eqn:filtpen}
  J_{\alpha,\sigma}[m,u;d]=\frac{1}{2}(\|K[u]F[m]w-d\|^2
  +\alpha^2\|tu\|^2 + \sigma^2\|u\|^2).
\end{equation}
As with all penalty methods, this definition involves a choice of
weight ($\alpha$). A second weight ($\sigma$) scales the norm-squared
of the filter, that is, a Tihonov regularization term, added for
technical reasons explained by \cite{Warner:16,Symes:24a}. Choice of
parameters is a critical step in the use of any penalty method. We
offer some remarks about this choice in the Discussion section.

The domain space (pairs of bulk modulus fields and adaptive filters)
is very high-dimensional, compared with the domain space for FWI (bulk
modulus fields alone, in the present context). It is possible to
optimize $J_{\alpha,\sigma}[m,u;d]$ by alternating updates of $m$ and
$u$  \cite[]{LiAlkhalifah:21}. However the {\em variable projection} \cite[]{GolubPereyra:73,GolubPereyra:03}
reduction is generally much more efficient than the altrnating, or
coordinate search, approach. In this instance, variable projection
consists in minimizing $J_{\alpha,\sigma}[m,u;d]$ over $u$ (a
quadratic optimization) to produce an optimal choice
$u_{\alpha,\sigma}[m:d]$. The reduced objective is
\begin{equation}
  \label{eqn:redfiltpen}
  \tilde{J}_{\alpha,\sigma}[m;d]=\frac{1}{2}(\|K[u_{\alpha,\sigma}[m;d]]F[m]w-d\|^2
  +\alpha^2\|tu_{\alpha,\sigma}[m;d]\|^2 + \sigma^2\|u_{\alpha,\sigma}[m;d]\|^2).
\end{equation}
Note that like the FWI objective, $\tilde{J}_{\alpha,\sigma}[m;d]$
depends only on $m$, with $d$ as a parameter.

\cite{Symes:24a} explains the relation between the reduced MSWI
objective and travel time inversion. This relation follows from the
geometric asymptotics approximation to solutions of the point radiator
problem \ref{eqn:awe} \cite[]{Friedlander:75}. Presuming that the
coefficients $\kappa, \beta$ are smooth, there are smooth functions
$\tau[m](\bx_s,\bx)$ (travel time) and $a[m](\bx_s,\bx)$ (geometric
amplitude) depending on the model $m$, so that
\begin{equation}
  \label{eqn:pwa}
  p(\bx,t;\bx_s) = a[m](\bx,\bx_s)w(t-\tau[m](\bx,\bx_s),\bx_s) + ...
\end{equation}
The elided terms are smoother than $w$, and become small with the
dominant wavelength in $w$ (\cite{Symes:94a} explains the precise
meaning of this condition). Thus we can write
\[
  F[m]w(\bx_r,t;\bx_s) \approx
  a[m](\bx_r,\bx_s)w(t-\tau[m](\bx_r,\bx_s))
\]
Supposing that the data $d$ is noise-free,
\[
  d(\bx_r,t;\bx_s) = F[m^*]w(\bx_r,t;\bx_s) \approx
  a[m^*](\bx_r,\bx_s)w(t-\tau[m^*](\bx_r,\bx_s))
\]
it follows that
\begin{equation}
  \label{eqn:tomo}
  \lim_{\alpha \rightarrow 0}\frac{1}{\alpha^2} (\tilde{J}_{\alpha,\sigma}[m;d]
  -\tilde{J}_{0,\sigma}) \approx \sum_{\bx_s,\bx_r} \frac{a[m^*]^2}{a[m]^2} (\tau[m^*]-\tau[m])^2\|g_{\frac{a[m^*]}{a[m]}\sigma}\|^2
\end{equation}
Here $g_{\sigma}$ is an approximate delta. Its Fourier transform of
is
\begin{equation}
  \label{eqn:gsig}
  \hat{g}_{\sigma} = \frac{a[m^*]^2 |\hat{w}|^2}{a[m^*]^2 |\hat{w}|^2
    + \sigma^2}
\end{equation}
whence $g_{\sigma}(\bx_s,\bx_r,t) \rightarrow \delta(t)$ in
the sense of distributions as $\sigma \rightarrow 0$.

If $\alpha$ and $\sigma$ are sufficiently small, then
$\tilde{J}_{\alpha,\sigma}$ is close to the right-hand side, which is bounded above and below by multiples of the the mean square travel time error between $m$ and
$m^*$.  This is the relation mentioned in the introduction. 
It does not
show that the only stationary points are those of the mean square
travel time error, even approximately. However if the amplitudes are
relatively insensive to changes in model, as is true if the
source-receiver distance is well away from developing multiple
arrivals, then the difference of stationary points gets small as $m$
approaches a stationary point of $\tilde{J}_{\alpha,\sigma}$. So we
would expect minimization $\tilde{J}_{\alpha,\sigma}$ to produce a
model that closely matches the travel times inherent in the data
($\tau[m^*]$ in the notation used here).

AWI adds one more feature, namely scaling by the trace norm of the
filter. This normalization can be interpreted as a choice of weighted
norm on the space of adaptive filters. It leads to a very similar
relation to \ref{eqn:tomo} for the AWI penalty function, but without
the amplitude factors: that is, the right-hand side of \ref{eqn:tomo}
is {\em just} the mean-square traveltime error. See \cite{Symes:24a}
for details.

The reduced adaptive filter $u_{\alpha,\sigma}[m;d]$ is the solution
of the {\em normal equation}
\begin{equation}
  \label{eqn:normal}
  (S[m]^TS[m] + \alpha^2 t^2 + \sigma^2 I)u = S[m]^Td,
\end{equation}
in which $S[m]u = K[u]F[m]w$. This positive definite symmetric linear
system may be solved by various efficient numerical methods. Having
computed $u_{\alpha,\sigma}[m;d]$ hence the value of 
$\tilde{J}_{\alpha,\sigma}[m;d]$, its (Euclidean) gradient is given by
\begin{equation}
  \label{eqn:gradredfiltpen}
  \nabla \tilde{J}_{\alpha,\sigma}[m;d] =
  D_m(F[m]w)^TK[u_{\alpha,\sigma}[m;d]]^T(K[u_{\alpha,\sigma}[m;d]]F[m]w-d)
\end{equation}
Apart from the appearance of the truncated filter operator
$K[u_{\alpha,\sigma}[m;d]]$, this is almost identical to the FWI
gradient \ref{eqn:fwigrad}. In particular, the last step in the
computation on the right hand side, the application of the adjoint
$m$-derivative of F, is exactly the same.

Use of a weighted norm in the model space goes exactly as before: with
weight operator $W$, the weighted gradient is
\begin{equation}
  \label{eqn:wgradredfiltpen}
  \nabla_W \tilde{J}_{\alpha,\sigma}[m;d] =
  W^{-1}D_m(F[m]w)^TK[u_{\alpha,\sigma}[m;d]]^T(K[u_{\alpha,\sigma}[m;d]]F[m]w-d)
\end{equation}

The gradient (or weighted gradient), together with the value, are the
inputs to first-order methods such as steepest descent and
Limited-Memory Broyden-Fletcher-Goldfarb-Shanno
iterations. Methods more closely related to Newton iteration require
more involved computations (see for instance \cite{Kaufman:75}.

@@

While the reduction in the (weighted) gradient norm indicates progress
towards a stationary point, it is not possible to claim that the final
estimate is in the vicinity of a local minimizer. The fit to data
obtained is not at all satisfactory. More discussion of this result
will be found below.

Now apply the same algorithm to the MSWI objective
$J_{\alpha,\sigma}$. Of course, to do so requires a choice of the
parameters $\alpha$ and $\sigma$. The purpose of the regularization
parameter $\sigma > 0$ is
simply to assure that the system \ref{eqn:normal} is positive
definite, so it can be chosen quite small - whatever that means. The
penalty parameter $\alpha > 0$ is the link between the filtered
residual and the spread in time of the filter: the performance of the
algorithm depends primarily on its choice. The next section will
discuss rational approaches to this choice. For now, we use the values
$\alpha=0.0001, \sigma=0.00001$ based on numerical exploration of the
objective.

The system \ref{eqn:normal} is far larger than is convenient to solve
by any variant of Gaussian Elimination, so iterative methods are 
necessary. We choose the Conjugate Gradient (CG.) method
\cite[]{Dan:67,Steihaug:83,NocedalWright}, and stop the
iteration by monitoring the reduction in length of the normal residual
(the gradient of $J_{\alpha,\sigma}[m,u;d]$ with respect to $u$). The
reduction threshhold $\rho$ is thus another necessary input parameter:
we choose $\rho = 0.001$ (note that unlike $\alpha$ and $\sigma$,
$\rho$ is dimensionless). We use the same inverse weight operator as
in the FWI example, namely a 10 $\times$ 10 moving average, repeated
once.

The initial bulk modulus field is once again homogeneous at 4 GPa. The
initial value of the reduced MSWI objective is $\approx 1.32 \times 10^{-2}$,
and the length of the (weighted) gradient is $\approx 6.37 \times
10^{-10}$. After 7 iterations, the normal residual reduction criterion
is satisfied. The value has decreased to $\approx 9.2 \times 10^{-4}$,
or a bit more than an order of magnitude reduction. The gradient
length is $\approx 4.33 \times 10^{-12}$, a reduction of more than two
orders of magnitude. The resulting bulk modulus is shown in Figure \ref{fig:mest11}.

\plot{mest11}{width=\textwidth}{a: Bulk modulus produced by 7
  steepest descent steps to minimize the reduced MSWI objective \ref{eqn:redfiltpen},
  using the data shown in \ref{fig:d11}.}

\multiplot{2}{sim11,resid11}{width=0.45\textwidth}{a: Data
  corresponding to MSWI result shown in Figure
  \ref{fig:mest11}. b: Residual or data error - difference between
  data shown in Figure \ref{fig:sim11} and target data
  \ref{fig:d11}. Note that the later signal in the central portion is
  still shifted, but by much less than in Figure \ref{fig:residfwi0}.}

Inspection of the residual plot \ref{fig:resid11} suggests that the
data predicted from the MSWI result is in many places within a
half-wavelength travel time shift of the data in Figure \ref{fig:d11}.
A half-wavelength is the conventional threshhold for FWI initial fit
error: below that it works, above that it doesn't. Thus it is tempting
to attempt FWI from using the MSWI result as initial estimate.

We applied 11 iterations of steepest descent to the FWI objective,
with initial estimate of $m$ equal to the final estimate from the MSWI
inversion just described (Figure \ref{fig:mest11}). The initial FWI
objective value was $\approx 4.3$, and the weighted gradient norm was
$\approx 2.15 \times 10^{-5}$.  After 11 steps of steepest descent,
the prescribed reduction ($10^{-2}$) of the weighted gradient norm was achieved,
and the objective value had decreased to $\approx 4.41 \times
10^{-2}$. The resulting bulk modulus field is depicted in Figure
\ref{fig:mestfwi}. The predicted data at this model, and the residual
data, are shown in Figures \ref{fig:resimfwi} and \ref{fig:residfwi}
respectively.

The FWI iteration starting at the MSWI result has effectively achieved
the global minimum of the FWI objective function: the data match is
very close. This type of result could perhaps have been achieved by
systematically increasing $\alpha$, as will be discussed in the next
section. However, the current result has a remarkable corollary,
provided that we are willing to assume that the intermediate values of
the FWI objective between steepest descent iterates are monotonically
decreasing. This condition could in principle be checked
computationally. However, for the moment let us accept that it is
so. Then we can conclude that the homogeneous model at which the first
FWI inversion was started must lie in a basin of attraction of a
non-global minimizer, since the value of the objective function at the
homogeneous model is actually {\em smaller} than that at the result of
MSWI inversion. However the descent sequence starting at that initial
estimate arrives at essentially zero residual. No descent path from
the homogeneous model can
